---
title: "Parallelization of SpatialDecon package"
author: "Marty Ross"
date: "2022-11-23"
output:
  rmarkdown::html_vignette: 
    toc: true
vignette: >
  %\VignetteIndexEntry{Use of SpatialDecon in a small GeoMx dataet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
p.caption {
  font-size: 1.5em;
}
</style>

### Installation

```{r installation, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if(!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("SpatialDecon")
```

### Data preparation

First, load the package:
```{r setup}
library(paraSpatialDecon)
suppressMessages(library(GeomxTools))
suppressMessages(library(tidyverse))
suppressMessages(library(kableExtra))
```

The data we use here is from the SpatialDecon package, the large kidney dataset. This parallelized implementation will take much the same inputs as the `spatialdecon` function: normalized and raw counts, as well as an annotation table describing each sample.

```{r loaddata , results='asis'}
data("kidney_gxt")
norm = as.matrix(assayDataElement(kidney_gxt, elt = "q_norm"))
raw = as.matrix(exprs(kidney_gxt))
annot = pData(kidney_gxt)

# missing ROI# field for some reason
annot <- as.data.frame(annot %>%
  group_by(`slide name`) %>%
  mutate(roi=str_pad(row_number(), 3, pad = "0")) %>%
  ungroup())
  
# New sample names
colnames(norm) = colnames(raw) = rownames(annot) = 
  paste0(annot$`slide name`, '_', annot$roi)

# Display sample types in a table
annot_tbl <- annot %>% 
  count(class, region) %>%
  group_by(class) %>%
  spread(region, n)

knitr::kable(annot_tbl, format = 'html', align = 'c') %>%
  kable_classic(full_width = F, html_font = "Cambria", 
                 bootstrap_options = "striped") %>%
  add_header_above(header = c('Sample Count' = 3), font_size = 16, bold = T) %>%
  row_spec(row = 0, bold = T, color = "white", background = "#888") %>%
  column_spec(column = 1, bold = T)

```

### Download a Profile Matrix 

You can download a profile matrix from NanoString's public repository on GitHub:

https://github.com/Nanostring-Biostats/CellProfileLibrary

For this vignette, we will load a kidney matrix derived from the HCA, which will also include associated metadata and a default binning of like cell types. The profile matrix is a cell-type-wise estimation of gene counts used in the deconvolution.

```{r get_nsdata}
profile_matrix <- SpatialDecon::download_profile_matrix(species = 'Human', age_group = 'Adult',
                                          matrixname = 'Kidney_HCA')
str(profile_matrix)
```

### Derive Assay Background
A modeled assay background can be generated by a `SpatialDecon` function, `derive_GeoMx_background`, which creates a probepool-wise matrix of assay noise. It will be used as an input for the decon function.

```{r bkgrnd}
bg = SpatialDecon::derive_GeoMx_background(norm = norm,
                                           probepool = rep(1, nrow(norm)),
                                           negnames = "NegProbe-WTX")
```

### Run paraspatialdecon
Now we'll run the `paraspatialdecon` function on our data.
For a profile matrix of middling  size, and 221 samples, this process might take a few minutes. Using 4 processors on my personal laptop, the parallelized process takes 2:26.

``` {r runparadc}
suppressMessages(require(SpatialDecon))
suppressMessages(require(tictoc))
tic()
fin_dclst <- paraspatialdecon(norm, raw, bg, profile_matrix, cellGroups,
                              cell_counts = annot$nuclei,
                              n_proc = 4)
toc()

```

### Run Standard spatialdecon Function for Comparison

Conversely, running on my laptop without parallelization, spatialdecon takes 3:48, roughly 2/3rd the time to run. One key aspect to understand about the `spatialdecon` algorithm is that run time scales roughly linearly by the number of samples run, but scales logarithmically by the number of distinct cell types being deconvolved.
Naturally, being able to distribute the job across more processors will enable greater time-savings

``` {r runspatdc}
# require(SpatialDecon)
require(tictoc)
tic()
orig_dclst <- spatialdecon(norm=norm, raw=raw, bg=bg, X=profile_matrix,
                           cellmerges=cellGroups,
                           cell_counts = annot$nuclei)
toc()

```

Now produce a binned proportional barplot of the results with annotation bars

``` {r basicbarplot, fig.height=7, fig.width=10, fig.cap = "Annotated", eval=T}
## Still have to subset and run another decon, because 221 samples is too many.

dc_barplot(fin_dclst, samp_notes = annot, annots = c('class','region'),
           proportional = TRUE, binned_graph = TRUE)



```
